# Assignment 4

## Problem 

According to one version of Moore’s law, the number of transistors on a state-of-the-art computer microprocessor roughly doubles every two years:

$$
C \approx \gamma 2^{A/2}
$$

where  

- $A$: the year number in which a microprocessor is introduced
- $C$: the transistor count of the microprocessor
- $\gamma$: a positive constant

Consider the **natural logarithm** of the transistor count: $\log C$

---

### a. Show that $\log C$ is roughly a simple linear regression and derive the coefficient of A

$$
\begin{aligned}
    \log C &\approx \log \gamma + \frac{A}{2} \log 2 \\
           &\approx \log \gamma + A \frac{1}{2} \log 2
\end{aligned}
$$

Given that the form of a simple linear regression is $y = \beta_1 + x \beta_2$ where $\beta_1$ and $\beta_2$ are the parameters, $\log C$ is roughly a simple linear regression with:  

- $y = \log C$
- $x = A$
- $\beta_1 = \log \gamma$
- $\beta_2 = \frac{1}{2} \log 2 = 0.3465736$ **(the coefficient of A)**

### b. Plot the data points as log transistor count versus year

```{r}
transistor <- read.csv("./Moores Law Data.csv", header = TRUE)
transistor$logC <- log(transistor$TransistorCount)
plot(
    x = transistor$Year, 
    y = transistor$logC, 
    main = "log(Transistor Count) vs. Year",
    ylab = "log(Transistor Count)",
    xlab = "Year"
)
```

### c. Consider a normal-theory simple linear regression model

$$
\log C_i \mid \beta, \sigma^2 A_i \sim indep.\;\; N(\beta_1 + \beta_2 (A_i - \bar A), \sigma^2)\quad i = 1,\dots, 235
$$

where  

- $\beta_1, \beta_2 \sim\quad iid. \; N(0, 1000^2)$
- $\sigma^2 \sim \text{Inv-gamma}(0.001, 0.001)$

#### i. List an appropriate JAGS model

```r
model {
    for (i in 1:length(Year)) {
        logC[i] ~ dnorm(beta_1 + beta_2 * (Year[i] - AvgYear[i]), tau)
    }
    beta_1 ~ dnorm(0, 1/1000000)
    beta_2 ~ dnorm(0, 1/1000000)
    tau ~ dgamma(0.001, 0.001)
    sigma.2 <- 1 / tau 
}
```

---

Now, run the model with overdispersed starting points, check convergence, and monitor $\beta_1$, $\beta_2$, and $\sigma^2$ for at least $2000$ iterations after burn-in.  

```{r}
transistor$AvgYear = mean(transistor$Year)
init.vals <- list(
    list(beta_1=10000, beta_2=10000, tau=0.00001),
    list(beta_1=-10000, beta_2=-10000, tau=0.00001),
    list(beta_1=10000, beta_2=10000, tau=0.01),
    list(beta_1=-10000, beta_2=-10000, tau=0.01)
)

library(rjags)
ml <- jags.model("./regression.bug", transistor, init.vals, n.chains=4, n.adapt=1000)
update(ml, 2500)
x <- coda.samples(ml, c("beta_1", "beta_2", "sigma.2"), n.iter=5000)

# verify convergence
gelman.diag(x, autoburnin=FALSE)
```

- As shown with the Gelman-Rubin convergence diagnostic, PSRF values equal to 1 for all parameters, indicating that MCMC chains have converged.

#### ii. List summary for $\beta_1$, $\beta_2$, and $\sigma^2$ 

```{r}
summary(x)
```

---

#### iii. Give the approximate posterior **mean** and 95% **central posterior interval** for the slope

- Slope is the statistics related to $\beta_2$

##### Slope Mean

```{r}
summary(x)$statistics["beta_2", "Mean"]
```

##### Slope 95% Central Posterior Interval

```{r}
x_mat = as.matrix(x)
quantile(x_mat[,"beta_2"], c(0.025, 0.975))
```

- No, the coefficient of A does not fall within the interval. It goes slightly beyond the upper end, which suggests that the estimated growth in transistor count is slightly lower than Moore's law.

---

#### iv. Give the approximate posterior **mean** and 95% **central posterior interval** for the error variance

##### Slope Mean

```{r}
summary(x)$statistics["sigma.2", "Mean"]
```

##### Slope 95% Central Posterior Interval

```{r}
quantile(x_mat[,"sigma.2"], c(0.025, 0.975))
```


### d. Predict the transistor count

#### i. Update JAGS model to answer the subsequent questions

```r
model {
    for (i in 1:length(Year)) {
        logC[i] ~ dnorm(beta_1 + beta_2 * (Year[i] - AvgYear[i]), tau)
    }
    beta_1 ~ dnorm(0, 1/1000000)
    beta_2 ~ dnorm(0, 1/1000000)
    tau ~ dgamma(0.001, 0.001)
    sigma.2 <- 1 / tau 

    mu_2026 <- beta_1 + beta_2 * (2026 - AvgYear[1])
    logC_2026 ~ dnorm(mu_2026, tau)
    C_2026 <- exp(logC_2026)
    C_2026_billions <- C_2026 / 1000000000

    invention_year <- AvgYear[1] - beta_1 / beta_2
}
```


```{r}
ml <- jags.model("./regression_d.bug", transistor, init.vals, n.chains=4, n.adapt=1000)
update(ml, 2500)
x <- coda.samples(ml, c("C_2026_billions", "invention_year"), n.iter=5000)

# verify convergence
gelman.diag(x, autoburnin=FALSE)
```


---

#### ii. List summary 

```{r}
summary(x)
```

---

#### iii. Give an approximate 95% central posterior predictive interval for the transistor count

```{r}
x_mat <- as.matrix(x)
quantile(x_mat[,"C_2026_billions"], c(0.025, 0.975))
```

---

#### iv. Explain the following formula and give the 95% central posterior interval 

$$
\bar A - \frac{\beta_1}{\beta_2}
$$

##### Proof

- When the transistor was invented, $C = 1$, which leads to $\log C = 0$
- Therefore,

$$
\begin{align*}
&\beta_1 + \beta_2 * (A - \bar A) = 0 \\
&\beta_2 A = \beta_2 \bar A - \beta_1 \\
&A = \bar A - \frac{\beta_1}{\beta_2}
\end{align*}
$$

##### 95% central posterior interval

```{r}
quantile(x_mat[,"invention_year"], c(0.025, 0.975))
```

##### Actual year

- Transistor was invented in [1947](https://en.wikipedia.org/wiki/History_of_the_transistor), which falls within the 95% central posterior interval


### e. Check for outliers

$$
T(y, X, \theta) = \max_i |\epsilon_i / \sigma|
$$

```{r}
ml <- jags.model("./regression.bug", transistor, init.vals, n.chains=4, n.adapt=1000)
update(ml, 2500)
x <- coda.samples(ml, c("beta_1", "beta_2", "sigma.2"), n.iter=5000)
x_mat <- as.matrix(x)
```

#### i. Compute simulated standardized error vectors $\epsilon / \sigma$

```{r}
n <- nrow(x_mat)
K <- length(transistor$logC)

eps_std <- matrix(NA, nrow = n, ncol = K)
for (s in 1:n) {
    beta1 <- x_mat[s, "beta_1"]
    beta2 <- x_mat[s, "beta_2"]
    sigma <- sqrt(x_mat[s, "sigma.2"])

    mu <- beta1 + beta2 * (transistor$Year - transistor$AvgYear)
    eps <- transistor$logC - mu
    eps_std[s,] <- eps / sigma
}
```

---

#### ii. Compute simulated replicate standardized errors $\epsilon^{rep} / \sigma$

```{r}
eps_rep_std <- matrix(NA, nrow = n, ncol = K)
for (s in 1:n) {
    beta1 <- x_mat[s, "beta_1"]
    beta2 <- x_mat[s, "beta_2"]
    sigma <- sqrt(x_mat[s, "sigma.2"])

    mu <- beta1 + beta2 * (transistor$Year - transistor$AvgYear)
    y_rep <- rnorm(K, mean = mu, sd = sigma)
    eps_rep_std[s,] <- (y_rep - mu) / sigma
}
```

---

#### iii. Compute $T(y, X, \theta)$ and $T(y^{rep}, X, \theta)$

##### $T(y, X, \theta)$

```{r}
T <- apply(abs(eps_std), 1, max)
```

##### $T(y^{rep}, X, \theta)$

```{r}
T_rep <- apply(abs(eps_rep_std), 1, max)
```


#### iv. Plot $T^{rep}$ vs $T$

```{r}
plot(
    T, 
    T_rep,
    xlab = "T(y, X, theta)",
    ylab = "T(y_rep, X, theta)",
    main = "Posterior Predictive Check"
)
abline(0, 1, col = "red")
```

#### Calculate the p-value

```{r}
mean(T_rep >= T)
```

- As the $\text{p-value}$ equals $0$, the observed extremes are always greater than or equal to the posterior model extremes. Therefore, we can infer that there must be an outlier in the observed data.

#### Name the microprocessor that appears to be the most extreme outlier

```{r}
x_stats = summary(x)$statistics
beta1_hat <- x_stats["beta_1", "Mean"]
beta2_hat <- x_stats["beta_2", "Mean"]
sigma_hat <- sqrt(x_stats["sigma.2", "Mean"])

mu_hat <- beta1_hat + beta2_hat * (transistor$Year - transistor$AvgYear)
std_resid <- abs((transistor$logC - mu_hat) / sigma_hat)

transistor[which.max(std_resid), "Processor"]
```
