---
format: 
  pdf:
    documentclass: scrartcl
    papersize: letter
    mainfont: "Times New Roman"
---

# Assignment 3

## Problem 1

### a. Simulates from the posteriors for $\mu$ and $\sigma^2$ and produce autocorrelation plots for their respective successive variates

```{r}
source("./Flint Gibbs.r", echo = FALSE)
acf(mu.sim)
acf(sigma.2.sim)
```

### b. Metropolis sampler

#### i. Find a value of `rho` that gives an average acceptance rate of about $0.35$

- With the `rho = 0.0308` the acceptance rate is around $0.35$

```{r}
rho <- 0.0308
source("./Flint Metropolis.r", echo = FALSE)
mean(accept.prob)
```

#### ii. Plot the successive autocorrelation plots for $\mu$ and $\sigma^2$

```{r}
acf(mu.sim)
acf(sigma.2.sim)
```

### c. Which method exhibited faster mixing

- **The Gibbs Sampler** exhibited faster mixing as the autocorrelation drops within the significance bounds faster compared to the Metropolis Sampler.


## Problem 2

### a. Use the model in `polls20161.bug` for the following questions

#### i. 

```{r}
d <- read.table("./Polls 2016.txt", header=TRUE)

# variable init
d$sigma <- d$ME / 2

# JAGS
library(rjags)
initial.vals <- list(
    list(mu=100, tau=0.01),
    list(mu=100, tau=100),
    list(mu=-100, tau=0.01),
    list(mu=-100, tau=100)
)

ml <- jags.model(
    "./Polls 2016.bug", 
    d, 
    initial.vals, 
    n.chains=4, 
    n.adapt=1000
)
```

#### ii. 

```{r}
update(ml, 2500) # burn-in

x <- coda.samples(ml, c("mu", "tau"), n.iter=5000)
```

#### iii.

- Based on the trace of `mu` and `tau`, there is no convergence problem.

```{r}
plot(x, smooth=FALSE)
```

#### iv.

- Based on the result of the Gelman-Rubin statistics, both $\mu$ and $\tau$ have the point estimation around $1.00$, which indicates a close-to-perfect convergence.

```{r}
gelman.diag(x, autoburnin=FALSE)
```


#### v. 

```{r}
# extract the first chain from x
first <- x[[1]]
autocorr.plot(first)
```

#### vi.

- Both are considered adequate as both of them exceed the suggested minimum value of $400$

```{r}
effectiveSize(x)
```

### b. A new model that uses an almost flat prior for `tau` on the **log** scale 

#### i. 

```r
model {

  for (j in 1:length(y)) {
    y[j] ~ dnorm(theta[j], 1/sigma[j]^2)
    theta[j] ~ dnorm(mu, 1/tau^2)
  }

  mu ~ dunif(-1000,1000)
  logtau ~ dunif(-100,100)
  tau <- exp(logtau)

}
```

#### ii.

```{r}
initial.vals <- list(
    list(mu=100, logtau=log(100)),
    list(mu=100, logtau=log(0.01)),
    list(mu=-100, logtau=log(100)),
    list(mu=-100, logtau=log(0.01))
)

ml <- jags.model(
    "./Polls 2016_2.bug",
    d, 
    initial.vals, 
    n.chains=4, 
    n.adapt=1000
)
```


#### iii.

```{r}
update(ml, 2500) # burn-in
x <- coda.samples(ml, c("mu", "tau"), n.iter=5000)
```


#### iv.

- Based on the trace of `mu` and `tau`, there is a convergence problem as the mixing of both variables are poor.
    - `mu`: The chains remain in separate regions.
    - `tau`: Most values are close to 0, with occasional spikes.  

```{r}
plot(x, smooth=FALSE)
```

#### v.

- Based on the result of the Gelman-Rubin statistics, both $\mu$ and $\tau$ have the point estimation greater than $1.1$, which indicates poor convergence.

```{r}
gelman.diag(x, autoburnin=FALSE)
```

#### vi.

- Based on the autocorrelation plots, the speed of mixing of both `mu` and `tau` is rather slow.

```{r}
autocorr.plot(x[[1]])
```

#### vii.

- The problem with this model is that using an imporper flat prior on $log \tau$ implies $p(\tau) \propto \frac{1}{\tau}$, which is itself improper. This can lead to an improper posterior, which is not a valid probability. Such a model can cause poor convergence behavior like the statistical results above.
